\plaintitle

### 2025-06-04 docs/style/article-body.tex
final version?

True random number generation is indispensable for a wide range of cryptographic and distributed computing applications, yet traditional pseudo-random number generators remain deterministic, and thus vulnerable to state‐recovery attacks. Quantum random number generation (QRNG) offers a robust alternative by harnessing inherent quantum fluctuations -- in particular, optical QRNG (OQRNG) based on shot noise provides an accessible source of entropy from which true random numbers can be generated. However, existing OQRNG solutions typically rely on expensive or bespoke hardware and defer post‐processing to host computers, limiting portability and integration. In this work, we investigate the feasibility of performing Toeplitz randomness extraction directly on resource constrained microcontrollers, thus creating a fully self‐contained QRNG system. We address two primary research areas: (RA1) developing an implementation of Toeplitz extraction that minimizes processing time on embedded hardware, and (RA2) ensuring that the extraction algorithm does not become the throughput bottleneck compared to ADC conversion or USB transfer speeds.

Our results demonstrate that, by eliminating dynamic allocations and leveraging bit‐parallel techniques, Toeplitz extraction on 64‐bit inputs can be executed in approximately \(0.05 \mu s\) on the Teensy 4.1 -- well below the \(2.667 \mu s\) threshold imposed by a 12‐bit, 2 MSPS ADC. Even accounting for variance in individual measurements, these speeds confirm that the algorithmic implementation does not constrain overall throughput. On the Pico 2, the same implementations yield sub‐microsecond performance, albeit with a wider variance. However, our most efficient implementations are limited to 64‐bit inputs -- extending to larger input bit sizes incurs significant overhead due to the absence of wider integer types in typical MCU toolchains.

@@

The optical signal output by the device described in \cite{Clason2023} needs to be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). Some post-processing of the said raw bits has to be performed in order to ensure that the bits are workable, and to remove potential deterministic patterns from the data. One method for this post-processing we will explore in this work is Toeplitz extraction \cite{toeplitz}, typically performed on the host computer utilizing the randomly generated numbers. The post-processing finally yields a random number. This relationship can be seen in Figure \ref{fig:linear-system}.

@@

In our work, we intend to use Teensy 4.1\footnote{\url{https://www.pjrc.com/store/teensy41.html}} based on the ARM Cortex-M7 pictured to the right in Figure \cite{fig:mcu}. This MCU is especially suitable for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and Digital Signal Processing (\emph{DSP}) capabilities. The floating-point unit\footnote{\url{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}} (\emph{FPU}) and Single Instruction, Multiple Data (\emph{SIMD}) style DSP instructions improve how quickly it can perform bitwise and arithmetic tasks, which is crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations -- making them highly effective for Toeplitz matrix-vector multiplications.

In order to evaluate how efficent our implementation can become, our aim is to try our implementation on other MCU with different level of power and hardware support. Whereas Teensy 4.1, operating at \texttt{600\ MHz}, is our primary development platform which we will evaluate closely, we aim to run our implementations on the Raspberry Pi Pico 2\footnote{\url{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}}, operating at \texttt{150\ MHz} (\emph{pictured to the right in Figure \cite{fig:mcu}}). Due to the lower computational power of this MCU, there may be significant issues in utilizing this weaker model, yet it is significantly cheaper and easier to access. Testing will consist solely of running the implementation on these controllers and measuring execution speed and correctness of the output.

@@

A detailed account of the inner workings of Toeplitz extraction can be found in the work of Chouhan et al. \cite{toeplitz-desc}. This work focuses on implementing Toeplitz extraction on field-programmable gate-arrays (FPGA), but some specific details can be derived from their work. As these authors describe, Toeplitz extraction is a strong contender for our work due to a lower computational complexity than other alternatives, as well as a relatively easy algorithm to use. This extraction utilizes either matrix multiplication or hashing between a pseudo-random seed and the raw data provided from a high-entropy source of randomness -- in our case, the OQRNG-device.

@@

\begin{algorithm}[ht] \caption{Toeplitz extraction}\label{alg:bit-conv}
\begin{algorithmic}[1] \REQUIRE \(x[0\,..\,n-1]\) \Comment{input bit array of
length \(n\)} \REQUIRE \(t[0\,..\,n+m-2]\) \Comment{seed matrix of length
\(n+m-1\)} \ENSURE \(y[0\,..\,m-1]\) \Comment{output bit array of length \(m\)}
\FOR{\(i = 0\) to \(m - 1\)} \STATE sum = 0 \FOR{\(j = 0\) to \(n - 1\)} \STATE
sum = sum + \(x[j] \* t[i + j]\) \ENDFOR \STATE \(y[i] =
\mathrm{sum}\;\bmod\;2\) \ENDFOR \RETURN \(y\) \end{algorithmic} \end{algorithm}

@@

Table \ref{tab:iter0} presents the average execution time of the naive implementation (iteration 0) across different input sizes for both the Teensy 4.1 and the Raspberry Pi Pico 2. The results reveal that the Teensy consistently achieves lower execution times compared to the Pico, with performance differences becoming more pronounced as thje bit size increases. This behaivor aligns with the expected difference in hardware specifications, such as CPU clock frequency and memory bandwidth.

>>>>

True random number generation is indispensable for a wide range of cryptographic and distributed computing applications, yet traditional pseudo-random number generators remain deterministic, and thus vulnerable to state‐recovery attacks. Quantum random number generation (\emph{QRNG}) offers a robust alternative by harnessing inherent quantum fluctuations -- in particular, optical QRNG (\emph{OQRNG}) based on shot noise provides an accessible source of entropy from which true random numbers can be generated. However, existing OQRNG solutions typically rely on expensive or bespoke hardware and defer post‐processing to host computers, limiting portability and integration. In this work, we investigate the feasibility of performing Toeplitz randomness extraction directly on resource constrained microcontrollers, thus creating a fully self‐contained QRNG system. We address two primary research areas: RA1 developing an implementation of Toeplitz extraction that minimizes processing time on embedded hardware, and RA2 ensuring that the extraction algorithm does not become the throughput bottleneck compared to ADC conversion or USB transfer speeds.

Our results demonstrate that, by eliminating dynamic allocations and leveraging bit parallel techniques, Toeplitz extraction on 64-bit inputs can be executed in approximately \(0.05 \mu s\) on the Teensy 4.1 -- well below the \(2.667 \mu s\) threshold imposed by a 12-bit, 2 MSPS ADC. Even accounting for variance in individual measurements, these speeds confirm that the algorithmic implementation does not constrain overall throughput. On the Pico 2, the same implementations yield sub‐microsecond performance, albeit with a wider variance. However, our most efficient implementations are limited to 64‐bit inputs -- extending to larger input bit sizes incurs significant overhead due to the absence of wider integer types in typical MCU toolchains.

@@

The optical signal output by the device described in \cite{Clason2023} needs to be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). Some post-processing of the said raw bits has to be performed in order to ensure that the bits are workable, and to remove potential deterministic patterns from the data. One method for this post-processing we will explore in this work is Toeplitz extraction \cite{toeplitz}, typically performed on the host computer utilizing the randomly generated numbers. The post-processing finally yields a random number. This relationship can be seen in fig.~\ref{fig:linear-system}.

@@

In our work, we intend to use Teensy 4.1\footnote{\url{https://www.pjrc.com/store/teensy41.html}} based on the ARM Cortex-M7 pictured to the right in fig.~\ref{fig:mcu}. This MCU is especially suitable for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and Digital Signal Processing (\emph{DSP}) capabilities. The floating-point unit\footnote{\url{https://www.sciencedirect.com/topics/computer-science/floating-point-unit}} (\emph{FPU}) and Single Instruction, Multiple Data (\emph{SIMD}) style DSP instructions improve how quickly it can perform bitwise and arithmetic tasks, which is crucial for quick Toeplitz extraction. SIMD-controlled DSP architectures, as described by Han et al. \cite{simd-dsp}, leverage parallel vectorized computation to accelerate matrix operations -- making them highly effective for Toeplitz matrix-vector multiplications.

In order to evaluate how efficient our implementation can become, our aim is to try our implementation on other MCU with different level of power and hardware support. Whereas Teensy 4.1, operating at \texttt{600\ MHz}, is our primary development platform which we will evaluate closely, we aim to run our implementations on the Raspberry Pi Pico 2\footnote{\url{https://datasheets.raspberrypi.com/pico/pico-2-product-brief.pdf}}, operating at \texttt{150\ MHz} (\_pictured to the right in fig.~\ref{fig:mcu_}). Due to the lower computational power of this MCU, there may be significant issues in utilizing this weaker model, yet it is significantly cheaper and easier to access. Testing will consist solely of running the implementation on these controllers and measuring execution speed and correctness of the output.

@@

A detailed account of the inner workings of Toeplitz extraction can be found in the work of Chouhan et al. \cite{toeplitz-desc}. This work focuses on implementing Toeplitz extraction on field-programmable gate-arrays (\emph{FPGA}), but some specific details can be derived from their work. As these authors describe, Toeplitz extraction is a strong contender for our work due to a lower computational complexity than other alternatives, as well as a relatively easy algorithm to use. This extraction utilizes either matrix multiplication or hashing between a pseudo-random seed and the raw data provided from a high-entropy source of randomness -- in our case, the OQRNG-device.

@@

\begin{algorithm}[ht] \caption{Toeplitz extraction}\label{alg:bit-conv}
\begin{algorithmic}[1] \REQUIRE \(k[0\,..\,n-1]\) \Comment{input bit array of
length \(n\)} \REQUIRE \(t[0\,..\,n+m-2]\) \Comment{seed matrix of length
\(n+m-1\)} \ENSURE \(y[0\,..\,m-1]\) \Comment{output bit array of length \(m\)}
\FOR{\(i = 0\) to \(m - 1\)} \STATE sum = 0 \FOR{\(j = 0\) to \(n - 1\)} \STATE
sum = sum + \(k[j] $\times$ t[i + j]\) \ENDFOR \STATE \(y[i] =
\mathrm{sum}\;\bmod\;2\) \ENDFOR \RETURN \(y\) \end{algorithmic} \end{algorithm}

@@

Table \ref{tab:iter0} presents the average execution time of the naive implementation (\emph{iteration 0}) across different input sizes for both the Teensy 4.1 and the Raspberry Pi Pico 2. The results reveal that the Teensy consistently achieves lower execution times compared to the Pico, with performance differences becoming more pronounced as thje bit size increases. This behaivor aligns with the expected difference in hardware specifications, such as CPU clock frequency and memory bandwidth.
